---
title: "Survival Models (MATH6143) Coursework"
author: "Student ID: 34273638"
geometry: margin = 2cm
output:
  pdf_document: default
header-includes:
  - \usepackage{wrapfig}
  - \usepackage{lipsum}
  - \usepackage{setspace}
  - \usepackage{titlesec}
  - \titlespacing{\title}{0pt}{\parskip}{-\parskip}
---
```{r,echo=F,message=F}
#Setting up enviroment
setwd("~/Desktop/Survival Models/CW/SM-CW")
lymp.data <- read.csv("lymphoma.csv",header=T)
duck.data <- read.table("duck.txt",header = T)
mortality.data <- read.csv("mortality.csv",header=T)
library(survival) #basic functions for survival modelling
library(survminer) # package to produce much nicer visualisations
library(tidyverse)
library(kableExtra)
library(readxl)
```
# Question 1:
\vspace{-5truemm}
## Part a:
Producing Kaplan Meier estimates of 38 patients with lymphocytic non-Hodgkins lymphoma who have been classified into 2 groups, symptomatic(n=10) and asymptomatic(n=28) gives the below plot.
```{r,include=F}
#Data explortation
lymp.data%>%
  group_by(symp)%>%
  summarise(n=n())

#Computing Kaplan Meier estimate of survivor function 
lymp.km <- survfit(Surv(time,event=cens) ~ symp, data=lymp.data)

summary(lymp.km)

#using log rank test
survdiff(Surv(time,event=cens) ~ symp, data=lymp.data)
```

```{r,echo=F,fig.align='center',fig.width=10,fig.height=4}
#Producing KM plot
plot(lymp.km,
     col=c("red","blue"),
     xlab = "Time (weeks)",
     ylab = "Survival probability",
     main = "Kaplan Meier survival estimates of patients \n with lymphocytic non-Hodgkins lymphoma")
legend("topright", legend = c("Asymptomatic", "Symptomatic"),
col = c("red", "blue"), lty = c(1, 1), cex = 0.8)
```
Viewing the plots it is clear that the asymptomatic group has a higher survival probability than the symptomatic group for all time, thus those in asymptomatic group have a longer survival time than those in the symptomatic group (on average). Furthermore, performing a log-rank test we attain a Chi-Squared statistic of 8.7 with 1 degree of freedom which corresponds to a significant p-value of **0.003** thus we reject the null hypothesis of no difference in survival between the groups. Therefore, we have evidence that there is a statistically significant difference of survival between the groups, backing up the analysis of the Kaplan Meier plot.

## Part b:
```{r,include=F}
#Finding CI (95%)
lymp.95 <- survfit(Surv(time,event=cens) ~ symp, data=lymp.data,conf.int=0.95)
summary(lymp.95)
#Using assumption that there are 52 weeks in a year, 6 year survival happens at week 312
#Reading off 95% confidence interval estimate at 312 weeks gives thew following 95% confidence intervals for the groups

#Asymptomatic (t>=301)
c(0.309, 0.690)
#Symptomatic (t>=281)
c(0.0156, 0.642)
```
For the asymptomatic and symptomatic groups the 95% confidence interval of the Kaplan Meier estimates of 6 year survival is [**0.309**, **0.690**] and [**0.0156**, **0.642**] respectively. 

## Part c:
```{r,include=F}
summary(lymp.km)
#For Asymptomatic time is some point greater than 301 weeks (study ended)
#For Symptomatic time is 276 weeks (drops from 0.3 to 0.2)
```
For the symptomatic group the estimated time the survival probability is below 30% is 276 weeks.

## Part d:
```{r,include=F}
#Part d
lymp.na <- survfit(Surv(time,event=cens) ~ symp, data=lymp.data,type = "fleming")


#Estimates are very similar, Nelson-Aalen estimates have survival portability shifted up (increased) slightly compared to Kaplan Meier
# estimates and this shift (increase in survival probability) is greater for the symptomatic group
summary(lymp.km)
summary(lymp.na)
#Survival estimates similar as d_{j}`/r_{j} small for all time for asymptomatic group (thus due to taylor approx they are approx equal)
#Survival estimates similar for symptomatic group too except at the end (due to small number at risk left) thus d_{j}`/r_{j} is not small
#so approximation is bad which is why the N-A estimates are more different for this group 

```

```{r,echo=F,fig.align='center'}
par(mfrow = c(1,2))
plot(lymp.na,
     col=c("red","blue"),
     xlab = "Time (weeks)",
     ylab = "Survival probability",
     main = "Nelson-Aalen survival estimates of patients \n with lymphocytic non-Hodgkins lymphoma",
     cex.main=0.8)
legend("topright", legend = c("Asymptomatic", "Symptomatic"),
col = c("red", "blue"), lty = c(1, 1), cex = 0.5)
plot(lymp.km,
     col=c("red","blue"),
     xlab = "Time (weeks)",
     ylab = "Survival probability",
     main = "Kaplan Meier survival estimates of patients \n with lymphocytic non-Hodgkins lymphoma",
     cex.main=0.8)
legend("topright", legend = c("Asymptomatic", "Symptomatic"),
col = c("red", "blue"), lty = c(1, 1), cex = 0.5)
```
Viewing the survival estimates via plots they are very similar. The Nelson-Aalen estimates have a survival probability shifted up (increased) slightly compared to the Kaplan Meier estimates and this shift (increase in survival probability) is greater for the symptomatic group. The survival estimates are similar for the asymptomatic group as the $\frac{d_{j}`}{r_{j}}$ (where $d_{j}$ is number of events at observed event time $t_{j}$ and $r_{j}$ is number at risk at $t_{j}$) is small for all observed event times $t_{j}$ with the largest value of $\frac{1}{13}$ at event time 301 weeks. Due to $\frac{d_{j}`}{r_{j}}$ being small for all $t_{j}$ the survival estimates are approximately equal ($\hat{S}_{NA}(t) \approx \hat{S}_{KM}(t)$) which can be proved by Taylor expansion. This is what also causes the change in estimates to be greater for the symptomatic group as the sample is smaller for this subgroup the value of $\frac{d_{j}`}{r_{j}}$ is larger with the largest value of 0.5 at event time 281 weeks. $\frac{d_{j}`}{r_{j}}$ is still small however as it is larger this causes the larger difference in survival estimates observed due to the approximation from Taylor expansion being less reliable.



## Part e:
```{r, include=F}
lymp.95.na <- survfit(Surv(time,event=cens) ~ symp, data=lymp.data,conf.int=0.95,type="fleming")
summary(lymp.95.na)
#95% CI of probability of 6-year (312 week) survival for each group

#Asymptomatic (t>=301)
c( 0.319, 0.697)
#Symptomatic (t>=281)
c(0.0340, 0.621)
```
For the asymptomatic and symptomatic groups the 95% confidence interval of the Nelson-Aalen estimates of 6 year survival is [**0.319**, **0.697**] and [**0.0340**, **0.621**] respectively. 

\newpage

# Question 2:
## Part a:
Fitting Weibull regression models to the duck survival time data I used AIC as my selection criteria and started with a model including all possible covariates and removed variables that lead to a new model with a smaller AIC. Starting with a hierarchical model with the three way interaction between age, weight and length (and thus all lower order terms), applying my model selection criteria, my final model was the null model with an AIC of **209.97** compared to an AIC of **216.08** for the initial model. This final model implies that all the covariates (age, weight and length) have no effect on the survival of the ducks. Moreover, creating a model with the covarites age, weight and length all are insignificant (had a p-value greater than 0.05). Thus, using the data provided I conclude that the survival of the ducks does not depend on explanatory variables age, weight and length.

```{r, include=F}
#Using AIC (information criteria for model selection we start with full model and perform backward selection)
duck.wb.all <- survreg(Surv(Time,event=Observed) ~ Age*Weight*Length, data = duck.data)
final.model.wb.1 <- step(duck.wb.all,data=duck.data,direction = "backward",trace = T)
AIC(final.model.wb.1)
#216.0846
#Note that the backward selection gets 'stuck' as all 2 way interactions but not 3 way has higher AIC (even though lower order models have lower AIC)

#`unsticking` model selection
duck.wb.2int <- survreg(Surv(Time,event=Observed) ~ Age*Weight+Weight*Length+Age*Length, data = duck.data)
final.model.wb.2 <- step(duck.wb.2int,data=duck.data,direction = "backward",trace = T)
AIC(final.model.wb.2)
#209.9668 - has lower AIC
final.model.wb.2$coefficients #final model is null model (only includes intercept)
final.wb <- final.model.wb.2

duck.wb.simple <- survreg(Surv(Time,event=Observed) ~ Age+Weight+Length, data = duck.data)
AIC(duck.wb.simple)
#Note that no interaction model has AIC of 213.5725
summary(duck.wb.simple)
```
## Part b:
A Weibull model assumes that the change in hazard rate is linear with time. MAYBE TALK ABOUT THIS ASSUMPTION?. Comparing the Weibull estimate with the Kaplan Meier estimate of survival gives the below plot. The plot shows that the Weibull estimate is consistent with the Kaplan Meier estimate thus the Weibull distribution is a reasonable model. (say somethign about KM being non-parametric thus can "trust")
```{r,echo=F,fig.align='center',fig.width=10,fig.height=5}
#Part b - Is Weibull distribution a reasonable model?
#Note that Weibull model assumes that hazard rate changes linearly across time (change in hazard is linear)
#This seem reasonable (ducks can live up to 20 years) so no change in there survival for 63 days for 1 year old for 0 year old may expect some change but
#as insignificant from year 1 no issue with this assumption
#Use comparison to Kaplan Meier plot for this

duck.km <- survfit(Surv(Time,event=Observed) ~ 1, data=duck.data)
plot(duck.km, ylab= "Survival function", xlab="Time (days)", col = "red",conf.int = F,
     main = "Estimated Kaplan Meier and Weibull survival curves \n for duck survival data")

alpha <- 1/final.wb$scale
beta_0 <- exp(-final.wb$coefficients[1])
curve(expr = exp(-(beta_0*x)^alpha), from = 0, to = 63, add = TRUE, col="blue")
legend("topright", legend = c("Kaplan Meier", "Weibull"),
col = c("red", "blue"), lty = c(1, 1), cex = 0.8)
#Looking at the curve it looks like a good approximation of the Kaplan Meier estimate
#Thus Weibull distribution is a reasonable model 
```

```{r, include=F}
duck.age.km <- survfit(Surv(Time,event=Observed) ~ Age, data=duck.data)
plot(duck.age.km, col = c("red", "blue"),
     xlab = "Time (days)",
     ylab = "Survival probability",
     main = "Kaplan Meier survival estimates of ducks split by age")
legend("topright", legend = c("Age: 0", "Age: 1"),
col = c("red", "blue"), lty = c(1, 1), cex = 0.8)
#Furthermore looking at the 2 Age groups for Kaplan Meier as they cross implies no
#significant difference in survival between 2 groups - as the insignificant p values
#was saying for Weibull regression models
```

## Part c:
Performing the same model selection criteria using the Cox proportional hazards model starting with the hierarchical model with the three way interaction between age, weight and length (and thus all lower order terms) my final model is the null model with a AIC of **128.00** compared to the initial models AIC of **135.44**. Thus, using my final model none of the covariates age, weight and length affect the survival of the ducks and when fitting a model with all 3 covariates, none are significant at $\alpha=0.05$. Therefore the survival of the ducks does not depend on its age, weight or length using the provided data.
```{r, include=F}
#Fitting cox models
duck.cox.all <- coxph(Surv(Time,event=Observed)~Age*Weight*Length,data=duck.data)
AIC(duck.cox.all)
duck.cox.final.1 <- step(duck.cox.all,data=duck.data,direction = "backward",trace = T)
AIC(duck.cox.final.1)

duck.cox.int2 <- coxph(Surv(Time,event=Observed) ~ Age*Weight+Weight*Length+Age*Length, data = duck.data)
duck.cox.final.2 <- step(duck.cox.int2,data=duck.data,direction = "backward",trace = T)
AIC(duck.cox.final.2)
final.cox <- duck.cox.final.2

summary(coxph(Surv(Time,event=Observed) ~ Age+Weight+Length, data = duck.data))
```

## Part d:
Plotting the estimated survivor function for a one year old duck with weight 1000g and length 250cm using my prefered models for part a and c is the survivor function for any duck as both models don't include the covariate age, weight or length. The corresponding plot is below.
```{r,echo=F,fig.align='center',fig.width=10,fig.height=5}
plot(survfit(final.cox),conf.int = F,col="red",
     xlab="Time (days)",
     ylab="Survival probabilty",
     main = "Estimated survival of ducks using \n Weibull and Cox regression")
curve(expr = exp(-(beta_0*x)^alpha), from = 0, to = 63, add = TRUE, col="blue")
legend("topright", legend = c("Cox", "Weibull"),
col = c("red", "blue"), lty = c(1, 1), cex = 0.8)
```
# Question 3:
## Part a:
```{r,include=F}
## Question 3 
mortality.data <- read.csv("mortality.csv",header=T)
#Part a - calculating crude central mortality rates
mortality.data<- mortality.data%>%
  mutate(cc.mort.rate.male=Male.deaths/Male.exposed,
         cc.mort.rate.female=Female.deaths/Female.exposed)%>%
  rename("Age"="Age..x.")

plot.mort.data.female <- mortality.data%>%
  dplyr::select(Age,cc.mort.rate.female)%>%
  mutate(gender="Female")%>%
  rename("cc.mort.rate"="cc.mort.rate.female")

plot.mort.data.male <- mortality.data%>%
  dplyr::select(Age,cc.mort.rate.male)%>%
  mutate(gender="Male")%>%
  rename("cc.mort.rate"="cc.mort.rate.male")

plot.mort.data <- rbind(plot.mort.data.female,plot.mort.data.male)%>%
  mutate(log.cc.mort.rate=log(cc.mort.rate))
```

Calculating the crude central mortality rates ($m_{x}$) for male and female pensioners and plotting it on the log scale gives the below graph. In it we can see that $log(m_{x})$ for males is higher than that of females for all ages with the exception of ages 63, 95, 96, 102. Thus, overall this graph suggests that $m_{x}$ is higher for males than females.
```{r,echo=F,fig.align='center',fig.width=10,fig.height=4}
#Plotting
ggplot(aes(x=Age,y=log.cc.mort.rate,colour=gender),data=plot.mort.data)+
  geom_point()+
  xlab("Age (years)")+
  ylab("Estimated log crude central mortality rate")+
  ggtitle("Log crude central mortality rates for members of large pension scheme")
```

## Part b:
Calculating $q_{x}$ for males and female using both the assumption of constant force of mortality and uniform distribution of deaths within each year of age gives the corresponding life tables (with $l_{60}=100000$).
